---
title: "LPA Personality Religion"
subtitle: "NZAVS"
abstract: |
  XXXX
author: 
  - name: VS
    affiliation: XXX
    orcid_id: XXXXX
    corresponding: yes
  - name: Don E Davis
    affiliation: Georgia State University
    orcid_id: 0000-0003-3169-6576 
  - name: Ken Rice
    affiliation: Georgia State University 
  - name: Geoffrey Troughton
    affiliation: Victoria University of Wellington
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland
    orcid_id: 0000-0002-4064-8800
  - name: Daryl Van Tongeren
    affiliation: Hope College
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid_id: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
keywords:
  - Author order TBA.
execute:
  warning: false
  eval: false
  echo: false
  include: false
editor_options: 
  chunk_output_type: console
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse") # data wrangling
library("tidyLPA") # latent profile analysis
library("patchwork") # making useful graphs 
library("table1") # tables
library("ggraph") # graphs
library("correlation")
library("report")
library("here") # file organisation
library("parameters") # data dimension reduction
library("DRR") # data dimension reduct
library("nFactors")


# read libraries
source("/Users/joseph/GIT/templates/functions/libs2.R")

# read functions
source("/Users/joseph/GIT/templates/functions/funs.R")

# experimental functions (more functions)
source(
  "https://raw.githubusercontent.com/go-bayes/templates/main/functions/experimental_funs.R"
)


# read data/ set to path in your computer
pull_path <-
  fs::path_expand(
    "/Users/joseph/v-project\ Dropbox/data/current/nzavs_13_arrow"
  )

# for saving models. # set path fo your computer
push_mods <-
  fs::path_expand(
    "/Users/joseph/v-project\ Dropbox/data/nzvs_mods/00drafts/23-lpa-personality"
  )

# for sourcing functions mid-script (for experimental scripts only)
path_funs<- 
  fs::path_expand(
    "/Users/joseph/GIT/templates/functions/funs.R"
  )

# read data: note that you need use the arrow package in R
dat <- arrow::read_parquet(pull_path)


# check path:is this correct?  check so you know you are not overwriting other directors
push_mods


```

## Data processing

Here, we perform a Latent Profile Analysis on a NZAVS 2018-2019 (Wave 10) which is the largest sample frame.


Data wrangling

```{r}

dat

dat$religion_believe_cats
# value              label
# 1       God-Spirit
# 2    God-No_Spirit
# 3    Spirit-No_God
# 4 No_God-No_Spirit

dat$religion_believe_cats

dat_18 <- dat |>
 arrange(id, wave) |>
  mutate(religion_believe_cats = factor(religion_believe_cats,
                                        levels = c(1, 2, 3, 4),
                                        labels = c("God-Spirit", "God-No_Spirit", "Spirit-No_God", "No_God-No_Spirit"))) |> 
# dplyr::filter(id != 9630) %>% # problematic reports no income but works full time and owns home.
  mutate(urban = factor(
    ifelse(
      rural_gch2018 == "Medium Urban Accessibility" |
        # Define urban condition
        rural_gch2018 == "High Urban Accessibility",
      "urban",
      # Label 'urban' if condition is met
      "rural"  # Label 'rural' if condition is not met
    )
  )) |>
  # select variables
  # mutate(across(where(is.double), as.numeric)) |>
  mutate(male = as.numeric(male) - 1) |>
  rename(religion_religious = religious) |>  # religious yes/no
  mutate(religion_church_binary = ifelse(religion_church > 0, 1, 0)) |>
  mutate(religion_religious = as.integer(as.numeric(as.character(religion_religious)))) |> 
  mutate(religion_believe_god_not = abs(religion_believe_god - 1)) |> 
  mutate(
    believe_god_not_spirit = ifelse(religion_believe_god == 1 &
                                      religion_believe_spirit == 0, 1, 0),
    believe_god_and_spirit =  ifelse(religion_believe_god == 1 &
                                       religion_believe_spirit == 1, 1, 0),
    not_believe_god_not_spirit =  ifelse(religion_believe_god == 0 &
                                           religion_believe_spirit == 0, 1, 0),
    not_believe_god_believe_spirit =  ifelse(religion_believe_god == 0 &
                                               religion_believe_spirit == 1, 1, 0)) |>
  mutate(religion_church_binary = ifelse(religion_church > 0, 1, 0)) |>
  dplyr::select(
    "wave",
    "year_measured",
    "id",
    # "edu",
    "sample_origin_names_combined",
    # Sample origin names combined
    #"alert_level_combined_lead",  not needed because all receive all levels by the point the outcome is measured
    # covid alert levels -> 2019-2020
    "education_level_coarsen",
    # Ordinal-Rank 0-10 NZREG codes (with overseas school quals coded as Level 3, and all other ancillary categories coded as missing)  Combined highschool levels See:https://www.nzqa.govt.nz/assets/Studying-in-NZ/New-Zealand-Qualification-Framework/requirements-nzqf.pdf
    "male",
    # 0 = female, 0.5 = neither female nor male, 1 = male.
    "age",
    "born_nz",
    "hlth_disability",
    "religion_believe_cats",
    
    # value label 0    No 1   Yes
    "eth_cat",
    #factor(EthCat, labels = c("Euro", "Maori", "Pacific", "Asian")),
    "employed",
    # Are you currently employed? (this includes self-employment or casual work)
    # "gen_cohort",
    "household_inc",
    # Please estimate your total household income (before tax) for the last year.
    "nz_dep2018",
    # see nzavs materials
    "nzsei13",
    # see nzavs materials
    "partner",
    # 0 = no, 1 = yes
    "parent",
    # 0 = no, 1 = yes
    "pol_orient",
    #Please rate how politically liberal versus conservative you see yourself as being.
    "pol_wing",
    # Please rate how politically left-wing versus right-wing you see yourself as being.
    "urban",
    # see NZAVS,
    "have_siblings", #Do you have siblings?
    "total_siblings",# sum siblings
    "number_sisters_older", #How many older sisters do you have?   
    "number_sisters_younger", #	How many younger sisters do you have? 
    "number_brothers_older",#	How many older brothers do you have?
    "number_brothers_younger", #	How many older brothers do you have?
    "children_num", # How many children have you given birth to, fathered, or adopted?
    "hours_children", #Hours - Looking after children
    "hours_work",#Hours - Working in paid employment
    "hours_housework", # Hours - Housework/cooking
    "agreeableness",
    # Mini-IPIP6 Agreeableness (also modelled as empathy facet)
    # Sympathize with others' feelings.
    # Am not interested in other people's problems.
    # Feel others' emotions.
    # Am not really interested in others.
    "conscientiousness",
    # see mini ipip6
    # Get chores done right away.
    # Like order.
    # Make a mess of things.
    # Often forget to put things back in their proper place.
    "extraversion",
    # Mini-IPIP6 Extraversion
    # Am the life of the party.
    # Don't talk a lot.
    # Keep in the background.
    # Talk to a lot of different people at parties.
    "honesty_humility",
    # see mini ipip6
    # Would like to be seen driving around in a very expensive car.
    # Would get a lot of pleasure from owning expensive luxury goods.
    # Feel entitled to more of everything.
    # Deserve more things in life.
    "openness",
    # see mini ipip6
    # Have a vivid imagination.
    # Have difficulty understanding abstract ideas.
    # Do not have a good imagination.
    # Am not interested in abstract ideas.
    "neuroticism",
    # see mini ipip6
    # Have frequent mood swings.
    # Am relaxed most of the time.
    # Get upset easily.
    # Seldom feel blue.
    "modesty",
    # see mini ipip6
    # I want people to know that I am an important person of high status,
    # I am an ordinary person who is no better than others.
    # I wouldn’t want people to treat me as though I were superior to them.
    # I think that I am entitled to more respect than the average person is
    # "sdo",
    # "rwa",
    # "brk_relationship",
    # "began_relationship",
    "religion_religious",
    # Do you identify with a religion and/or spiritual group?
   # "religion_religious_not",  # reverse this indicator
    "religion_identification_level",
    #How important is your religion to how you see yourself?"
    "religion_church_binary",
  #  "religion_church_binary2",
    "religion_prayer",
    # How many times did you pray in the last week?
    "religion_scripture",
    # How many times did you read religious scripture in the last week?
    "religion_church2",
  #  # How many times did you attend a church or place of worship in the last month?
    "religion_believe_spirit",
    #Do you believe in some form of spirit or lifeforce?
    "religion_believe_spirit",  #inverse believe in god
    "religion_believe_god",
    #Do you believe in a God
    "religion_believe_god_not",  #inverse believe in god
    "religion_spiritual_identification",
    #w8,w10,w12-13 "I identify as a spiritual person."
    "religion_perceive_religious_discrim",
    #	I feel that I am often discriminated against because of my religious/spiritual beliefs.
    # "bigger_doms", #What religion or spiritual group?#  Not_Rel, Anglican , Buddist, Catholic , Christian_nfd, Christian_Others, Hindu, Jewish           Muslim, PresbyCongReform, TheOthers
    "w_gend_age_euro",
    # sample_weights.
    # Sometimes I can't sleep because of thinking about past wrongs I have suffered.//# I can usually forgive and forget when someone does me wrong.# I find myself regularly thinking about past times that I have been wronged.
    "gratitude",
    ## I have much in my life to be thankful for. # When I look at the world, I don’t see much to be grateful for. # I am grateful to a wide variety of peopl
    "modesty",
    # see above
    "vengeful_rumin",
    "charity_donate",
    #How much money have you donated to charity in the last year?
    "hours_charity",
    #,#Hours spent in activities/Hours spent … voluntary/charitable work
    "warm_asians",
    "warm_chinese",
    #"warm_disabled" ,  missing at time 0
    # begins w9
    "warm_immigrants",
    "warm_indians",
    "warm_elderly",
    # warm_lgbtq starts w12
    "warm_maori",
    "warm_mental_illness",
    "warm_muslims",
    "warm_nz_euro",
    "warm_overweight",
    "warm_pacific",
    "warm_refugees",
    "religion_perceive_religious_discrim",
    # "issue_same_sex_marriage", not in range
     "support", # three items as below
    # "support_help",
    # # 'There are people I can depend on to help me if I really need it.
    # "support_turnto",
    # # There is no one I can turn to for guidance in times of stress.
    # "support_rnoguidance",
    #There is no one I can turn to for guidance in times of stress.
    "family_time",
    "friends_time",
    "community_time",
    "family_money",
    "friends_money",
    "community_money",
     #Please estimate how much help you have received from the following sources in the last week?
    # Received help and support - hours
    # family
    # friends
    # others in my community
    # Received help and support - money
    # family
    # friends
    # others in my community
    # outcomewide, 
     "religion_religious",
    # Do you identify with a religion and/or spiritual group?
    "religion_identification_level",
    #How important is your religion to how you see yourself?"
    "religion_church_binary",
    "religion_prayer",
    # How many times did you pray in the last week?
    "religion_scripture",
    # How many times did you read religious scripture in the last week?
    "religion_church2",
    # How many times did you attend a church or place of worship in the last month?
    "religion_believe_spirit",
    #Do you believe in some form of spirit or lifeforce?
    "religion_believe_god",
    #Do you believe in a God
    "religion_spiritual_identification",
    #w8,w10,w12-13 "I identify as a spiritual person."
    "religion_perceive_religious_discrim",
    #	I feel that I am often discriminated against because of my religious/spiritual beliefs.
    # "bigger_doms", #What religion or spiritual group?#  Not_Rel, Anglican , Buddist, Catholic , Christian_nfd, Christian_Others, Hindu, Jewish           Muslim, PresbyCongReform, TheOthers
    # sample_weights
    "alcohol_frequency",
    #"How often do you have a drink containing alcohol?"
    "alcohol_intensity",
    # How many drinks containing alcohol do you have on a typical day when drinking?
    "hlth_bmi",
    # " What is your height? (metres)\nWhat is your weight? (kg)\nKg
    "hours_exercise",
    # Hours spent … exercising/physical activity
    # "sfhealth",
    "sfhealth_your_health",
    # "In general, would you say your health is...
    "sfhealth_get_sick_easier",
    #\nI seem to get sick a little easier than other people.
    "sfhealth_expect_worse_health",
    #\nI expect my health to get worse." ****
    "hlth_sleep_hours",
    #During the past month, on average, how many hours of actual sleep did you get per night?
    "smoker",
    #Do you currently smoke?
    "hlth_fatigue",
    #During the last 30 days, how often did.... you feel exhausted?
    "rumination",
    # During the last 30 days, how often did.... you have negative thoughts that repeated over and over?
    "kessler_depressed",
    #During the last 30 days, how often did.... you feel so depressed that nothing could cheer you up?
    "kessler_effort",
    #During the last 30 days, how often did.... you feel that everything was an effort?
    "kessler_hopeless",
    # During the last 30 days, how often did.... you feel hopeless?
    "kessler_nervous",
    #During the last 30 days, how often did.... you feel nervous?
    "kessler_restless",
    #During the last 30 days, how often did.... you feel restless or fidgety?
    "kessler_worthless",
    # During the last 30 days, how often did.... you feel worthless?
    "sexual_satisfaction",
    #  How satisfied are you with your sex life?
    "bodysat",
    ## Am satisfied with the appearance, size and shape of my body.
    "vengeful_rumin",
    # Sometimes I can't sleep because of thinking about past wrongs I have suffered.//# I can usually forgive and forget when someone does me wrong.# I find myself regularly thinking about past times that I have been wronged.
    "perfectionism",
    # # Doing my best never seems to be enough./# My performance rarely measures up to my standards.
    # I am hardly ever satisfied with my performance.
    "power_self_nocontrol",
    # I do not have enough power or control over\nimportant parts of my life.
    "power_others_control",
    # Other people have too much power or control over\nimportant parts of my life
    "self_esteem",
    "selfesteem_satself",
    #  On the whole am satisfied with myself.
    "selfesteem_postiveself",
    # Take a positive attitude toward myself
    "selfesteem_rfailure",
    # Am inclined to feel that I am a failure.
  #  "self_control",
    "self_control_have_lots",
    #In general, I have a lot of self-control.
    "self_control_wish_more_r",
    #I wish I had more self-discipline.(r)
    "emotion_regulation_out_control",
    # When I feel negative emotions, my emotions feel out of control. w10 - w13
    "emotion_regulation_hide_neg_emotions",
    # When I feel negative emotions, I suppress or hide my emotions. w10 - w13
    "emotion_regulation_change_thinking_to_calm",
    # When I feel negative emotions, I change the way I think to help me stay calm. w10 - w13
    # "emp_work_life_balance",# I have a good balance between work and other important things in my life. # not measured at baseline
   # "respect_self",  #If they knew me, most NZers would respect what I have accomplished in life. Missing at T12
    "gratitude",
    ## I have much in my life to be thankful for. # When I look at the world, I don’t see much to be grateful for. # I am grateful to a wide variety of people.
    "pwi_health",
    #Your health.
    "pwi_relationships",
    #Your personal relationships.
    "pwi_security",
    #Your future security.
    "pwi_standardliving",
    #Your standard of living.
    "lifesat",
    "lifesat_satlife",
    # I am satisfied with my life.
    "lifesat_ideal",
    # In most ways my life is close to ideal.
    "lifemeaning", # average meaning_purpose, meaning_sense
    "meaning_purpose",
    # My life has a clear sense of purpose.
    "meaning_sense",
    # I have a good sense of what makes my life meaningful.
    "permeability_individual",
    #I believe I am capable, as an individual\nof improving my status in society.
    "impermeability_group",
    #The current income gap between New Zealand Europeans and other ethnic groups would be very hard to change.
    "neighbourhood_community",
    #I feel a sense of community with others in my local neighbourhood.
    "support",
    "support_help",
    # 'There are people I can depend on to help me if I really need it.
    "support_turnto",
    # There is no one I can turn to for guidance in times of stress.
    "support_rnoguidance",
    #There is no one I can turn to for guidance in times of stress.
    "belong",
    "belong_accept",
    #Know that people in my life accept and value me.
    "belong_routsider",
    # Feel like an outsider.
    "belong_beliefs",
    # Know that people around me share my attitudes and beliefs.
    "charity_donate",
    #How much money have you donated to charity in the last year?
    "hours_charity",#,#Hours spent in activities/Hours spent … voluntary/charitable work
    # "nwi", # The economic situation in New Zealand./# The social conditions in New Zealand. # Business in New Zealand.
    "emp_job_sat", # How satisfied are you with your current job? #Eisenbarth, H., Hart, C. M., Zubielevitch, E., Keilor, T., Wilson, M. S., Bulbulia, J. A., Sibley, C. G., & Sedikides, C. (in press). Aspects of psychopathic personality relate to lower subjective and objective professional success. Personality and Individual Differences, 186, 111340.
    "emp_job_secure",  #only for employed people
    "emp_job_valued"
  )|> 
  dplyr::rename(sample_weights = w_gend_age_euro) |> 
  filter(year_measured ==1 & wave ==2018) |> 
  droplevels()

n_unique(dat_18$id) # 47948


df <- dat_18 |> 
  mutate(believe_god_not_spirit = ifelse(religion_believe_god == 1 &
                                      religion_believe_spirit == 0, 1, 0),
    believe_god_and_spirit =  ifelse(religion_believe_god == 1 &
                                       religion_believe_spirit == 1, 1, 0),
    not_believe_god_not_spirit =  ifelse(religion_believe_god == 0 &
                                           religion_believe_spirit == 0, 1, 0),
    not_believe_god_believe_spirit =   ifelse(religion_believe_god == 0 &
                                               religion_believe_spirit == 1, 1, 0)) |> 
  select(
    id,
    religion_believe_cats,
    believe_god_and_spirit,
    believe_god_not_spirit,
    not_believe_god_believe_spirit,
    not_believe_god_not_spirit,
    "agreeableness",
    # Mini-IPIP6 Agreeableness (also modelled as empathy facet)
    # Sympathize with others' feelings.
    # Am not interested in other people's problems.
    # Feel others' emotions.
    # Am not really interested in others.
    "conscientiousness",
    # see mini ipip6
    # Get chores done right away.
    # Like order.
    # Make a mess of things.
    # Often forget to put things back in their proper place.
    "extraversion",
    # Mini-IPIP6 Extraversion
    # Am the life of the party.
    # Don't talk a lot.
    # Keep in the background.
    # Talk to a lot of different people at parties.
    "honesty_humility",
    # see mini ipip6
    # Would like to be seen driving around in a very expensive car.
    # Would get a lot of pleasure from owning expensive luxury goods.
    # Feel entitled to more of everything.
    # Deserve more things in life.
    "openness",
    # see mini ipip6
    # Have a vivid imagination.
    # Have difficulty understanding abstract ideas.
    # Do not have a good imagination.
    # Am not interested in abstract ideas.
    "neuroticism",
    # see mini ipip6
    # Have frequent mood swings.
    # Am relaxed most of the time.
    # Get upset easily.
    # Seldom feel blue.
    "modesty"
    # see mini ipip6
    # I want people to know that I am an important person of high status,
    # I am an ordinary person who is no better than others.
    # I wouldn’t want people to treat me as though I were superior to them.
    # I think that I am entitled to more respect than the average person is
  ) |> 
  mutate(agreeableness_s = scale(agreeableness),
         conscientiousness_s = scale(conscientiousness),
         extraversion_s = scale(extraversion),
         honesty_humility_s = scale(honesty_humility),
         openness_s = scale(openness),
         neuroticism_s = scale(neuroticism),
         modesty_s = scale(modesty))
           

table(df$believe_god_and_spirit)
table(df$believe_god_not_spirit)
table(df$not_believe_god_believe_spirit)
table(df$not_believe_god_not_spirit)


## regressions:

parameters::parameters( glm(believe_god_and_spirit ~ agreeableness_s, family = binomial, data = df) )
parameters::parameters( glm(believe_god_and_spirit ~ conscientiousness_s, family = binomial, data = df) )
parameters::parameters( glm(believe_god_and_spirit ~ honesty_humility_s, family = binomial, data = df) )
parameters::parameters( glm(believe_god_and_spirit ~ openness_s, family = binomial, data = df) )
parameters::parameters( glm(believe_god_and_spirit ~ neuroticism_s, family = binomial, data = df) )
parameters::parameters( glm(believe_god_and_spirit ~ modesty_s, family = binomial, data = df) )

parameters::parameters( glm(not_believe_god_not_spirit ~ agreeableness_s, family = binomial, data = df) )
parameters::parameters( glm(not_believe_god_not_spirit ~ conscientiousness_s, family = binomial, data = df) )
parameters::parameters( glm(not_believe_god_not_spirit ~ honesty_humility_s, family = binomial, data = df) )
parameters::parameters( glm(not_believe_god_not_spirit ~ openness_s, family = binomial, data = df) )
parameters::parameters( glm(not_believe_god_not_spirit ~ neuroticism_s, family = binomial, data = df) )
parameters::parameters( glm(not_believe_god_not_spirit ~ modesty_s, family = binomial, data = df) )



df_believe_god_and_spirit <- df |> filter(believe_god_and_spirit == 1) |>
  select(
    agreeableness,
    conscientiousness,
    extraversion,
    honesty_humility,
    openness,
    neuroticism,
    modesty
  ) |>
  droplevels()

nrow(df_god_spirit)

df_believe_god_not_spirit <-  df |> filter(believe_god_not_spirit == 1) |>
  select(
    agreeableness,
    conscientiousness,
    extraversion,
    honesty_humility,
    openness,
    neuroticism,
    modesty
  ) |>
  droplevels()

nrow(df_believe_god_not_spirit)

df_not_believe_god_believe_spirit <-
  df |>  filter(not_believe_god_believe_spirit == 1) |>
  select(
    agreeableness,
    conscientiousness,
    extraversion,
    honesty_humility,
    openness,
    neuroticism,
    modesty
  ) |>
  droplevels()

nrow(df_not_believe_god_believe_spirit)

df_not_believe_god_not_spirit <-
  df |>  filter(not_believe_god_not_spirit == 1) |>
  select(
    agreeableness,
    conscientiousness,
    extraversion,
    honesty_humility,
    openness,
    neuroticism,
    modesty
  ) |>
  droplevels()
droplevels()


  nrow(df_not_believe_god_not_spirit)

```

Plot relationship of variables

```{r  cache = TRUE, layout="l-body-outset",  fig.height=8, fig.width=12}
colnames(df_god_spirit)

df_believe_god_not_spirit %>%
  dplyr::select(-id)%>%
  correlation::correlation(partial = FALSE, multilevel = FALSE ) %>%
  plot()
```

We create a simple data summary to ensure that all variables look OK. We can see that rats and wb are strongly anti-correlated, which provides a sanity check. 

```{r}
library(skimr)
table1::table1(~ agreeableness+
    conscientiousness+
    extraversion+
    honesty_humility+
    openness+
    neuroticism, data = df)
```

I find it easier to see what is going on in the data by inspecting histograms of the variables:

```{r layout="l-body-outset",  fig.height=8, fig.width=12}
skimr::skim(df, -id)
```



<!-- ```{r  cache = TRUE, layout="l-body-outset",  fig.height=8, fig.width=12} -->
<!-- sh_hist <- ggplot2::qplot(data = df, sh, geom = "histogram") + theme_classic() +  -->
<!--   labs(title = "Histogram of sh") -->
<!-- stht_hist <- ggplot2::qplot(data = df, stht, geom = "histogram") + theme_classic() +  -->
<!--   labs(title = "Histogram of stht") -->
<!-- satt_hist <- ggplot2::qplot(data = df, satt, geom = "histogram") + theme_classic() +  -->
<!--   labs(title = "Histogram of satt") -->

<!-- sh_hist/stht_hist/satt_hist + plot_annotation(title= "We might want to condsider whether stht is given sufficient variation", tag_levels = "a") -->
<!-- ``` -->

<!-- Here a table is more informative: -->

<!-- # ```{r   cache = TRUE} -->
<!-- # table1::table1(~ -->
<!-- #                  factor(sh) + -->
<!-- #                  factor(stht) + -->
<!-- #                  factor(satt), -->
<!-- #                data = df -->
<!-- #                ) -->
<!-- # ``` -->



## lpa_believe_god_not_spirit

```{r  cache = TRUE}
library(arm)
lpa_believe_god_not_spirit <- df_believe_god_not_spirit %>%
  # dplyr::mutate(
  #   anx_s = as.numeric(scale(anx)),
  #   cdct_s = as.numeric(scale(cdct))  ,
  #   sh_s = as.numeric(scale(sh)),
  #   stht_s = as.numeric(scale(stht)),
  #   wb_s = as.numeric(scale(wb)),
  #   satt_s = as.numeric(scale(satt)),
  #   rads_s = as.numeric(scale(rads))
  # ) |> 
  tidyLPA::single_imputation() %>%
  tidyLPA::estimate_profiles(2:4) # best = 2

#model
lpa_believe_god_not_spirit
plot_lpa_believe_god_not_spirit <- tidyLPA::plot_profiles( lpa_believe_god_not_spirit[[1]], add_line = TRUE) + ggtitle( "Believe God Not believe Spirit", subtitle = "type 1 = 30%; type 2: 70%")
plot_lpa_believe_god_not_spirit
```

```{r}
lpa_believe_god_not_spirit_2 <- tidyLPA::get_data(lpa_believe_god_not_spirit[[1]]) ## note we are counting up from three, hence `[[2]]`
lpa_believe_god_not_spirit_2 %>%
  dplyr::select(Class) %>%
  dplyr::mutate(class = as.factor(Class)) %>%
  dplyr::group_by(class) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(freq = n / sum(n))
```




## df_believe_god_and_spirit


```{r  cache = TRUE}
library(arm)
lpa_believe_god_and_spirit <- df_believe_god_and_spirit %>%
  # dplyr::mutate(
  #   anx_s = as.numeric(scale(anx)),
  #   cdct_s = as.numeric(scale(cdct))  ,
  #   sh_s = as.numeric(scale(sh)),
  #   stht_s = as.numeric(scale(stht)),
  #   wb_s = as.numeric(scale(wb)),
  #   satt_s = as.numeric(scale(satt)),
  #   rads_s = as.numeric(scale(rads))
  # ) |> 
  tidyLPA::single_imputation() %>%
  tidyLPA::estimate_profiles(2:4) # best = 2

#model

plot_lpa_believe_god_and_spirit<- tidyLPA::plot_profiles( lpa_believe_god_and_spirit[[1]], add_line = TRUE)+ ggtitle( "Believe God And Believe Spirit", subtitle = "type 1 = 24%; type 2: 76%")
plot_lpa_believe_god_and_spirit

```

```{r}
lpa_believe_god_and_spirit_2 <- tidyLPA::get_data(lpa_believe_god_and_spirit[[1]]) ## note we are counting up from three, hence `[[2]]`
lpa_believe_god_and_spirit_2 %>%
  dplyr::select(Class) %>%
  dplyr::mutate(class = as.factor(Class)) %>%
  dplyr::group_by(class) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(freq = n / sum(n))
```


### df_not_believe_god_believe_spirit


```{r  cache = TRUE}
library(arm)
lpa_not_believe_god_believe_spirit <- df_not_believe_god_believe_spirit %>%
  # dplyr::mutate(
  #   anx_s = as.numeric(scale(anx)),
  #   cdct_s = as.numeric(scale(cdct))  ,
  #   sh_s = as.numeric(scale(sh)),
  #   stht_s = as.numeric(scale(stht)),
  #   wb_s = as.numeric(scale(wb)),
  #   satt_s = as.numeric(scale(satt)),
  #   rads_s = as.numeric(scale(rads))
  # ) |> 
  tidyLPA::single_imputation() %>%
  tidyLPA::estimate_profiles(2:4) # best = 2

#model

plot_lpa_not_believe_god_believe_spirit<- tidyLPA::plot_profiles( lpa_not_believe_god_believe_spirit[[1]], add_line = TRUE)

plot_lpa_not_believe_god_believe_spirit

```

```{r}
lpa_believe_god_and_spirit_2 <- tidyLPA::get_data(lpa_believe_god_and_spirit[[1]]) ## note we are counting up from three, hence `[[2]]`
lpa_believe_god_and_spirit_2 %>%
  dplyr::select(Class) %>%
  dplyr::mutate(class = as.factor(Class)) %>%
  dplyr::group_by(class) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(freq = n / sum(n))
```



##     df_not_believe_god_not_spirit


```{r  cache = TRUE}
library(arm)
lpa_not_believe_god_not_believe_spirit <- df_not_believe_god_not_spirit %>%
  # dplyr::mutate(
  #   anx_s = as.numeric(scale(anx)),
  #   cdct_s = as.numeric(scale(cdct))  ,
  #   sh_s = as.numeric(scale(sh)),
  #   stht_s = as.numeric(scale(stht)),
  #   wb_s = as.numeric(scale(wb)),
  #   satt_s = as.numeric(scale(satt)),
  #   rads_s = as.numeric(scale(rads))
  # ) |> 
  tidyLPA::single_imputation() %>%
  tidyLPA::estimate_profiles(2:4) # best = 2

#model
lpa_not_believe_god_not_believe_spirit

plot_lpa_not_believe_god_not_believe_spirit<- tidyLPA::plot_profiles( lpa_not_believe_god_not_believe_spirit[[1]], add_line = TRUE) + ggtitle( "Not Believe God But Believe Spirit", subtitle = "type 1 = 23%; type 2: 77%")

plot_lpa_not_believe_god_not_believe_spirit

```

```{r}
lpa_not_believe_god_believe_spirit_2 <- tidyLPA::get_data(lpa_not_believe_god_believe_spirit[[1]])
lpa_not_believe_god_believe_spirit_2 %>%
  dplyr::select(Class) %>%
  dplyr::mutate(class = as.factor(Class)) %>%
  dplyr::group_by(class) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(freq = n / sum(n))
```




```{r}
plot_lpa_believe_god_not_spirit / plot_lpa_believe_god_and_spirit /plot_lpa_not_believe_god_believe_spirit/plot_lpa_not_believe_god_not_believe_spirit
```




## An "experimental" approach

An alternative approach might be to understand whether we can reduce features of the data and *then* apply LPA.`r knitr::asis_output("\U1F923")`


### Try a PCA

```{r}
ddf <- df %>%
  dplyr::select(-id)

# PCA
set.seed(12)
out <-parameters::reduce_parameters(ddf, method = "PCA", n = "max", distance = "euclidean")
head(out)
```

WARNING: I'm not sure it makes sense to perform an LPA on a PCA dataset, but here it goes

### Prepare the data 

```{r cache = TRUE}
# rename the columns
out1 <-  out %>%
  dplyr::rename(
    wbPCA =  "rads_-0.87/stht_-0.74/sh_-0.73/wb_0.72",
    sattPCA = "satt_0.65",
    cdctPCA = "cdct_-0.84",
    anxPCA = "anx_0.66"
  )
```

### LPA

```{r cache = TRUE}
## fit an LPA on the Principle Components
fit4 <- out1 %>%
 dplyr::select(wbPCA,
         sattPCA,
         cdctPCA,
         anxPCA) %>%
  tidyLPA::single_imputation() %>%
  tidyLPA::estimate_profiles(3:6)
```


```{r cache = TRUE, layout="l-body-outset",  fig.height=12, fig.width=12}
# Table of fits
tidyLPA::get_fit( fit4 )
```

### Results

A three class solution seems OK.  let's graph this one:


```{r  cache = TRUE, layout="l-body-outset",  fig.height=8, fig.width=12}
tidyLPA::plot_profiles( fit4, add_line = TRUE ) + 
  ylab("PCA units") + labs(title = "LPA of PCA: no warrenties")
```



Let's graph the four class solution on its own


```{r  cache = TRUE, layout="l-body-outset",  fig.height=8, fig.width=12}

tidyLPA::plot_profiles( fit4[[2]], add_line = TRUE ) + 
  ylab("PCA units") + labs(title = "LPA of PCA: no warrenties")
```



Calculate frequencies:




```{r cache = TRUE}
d5 <- tidyLPA::get_data( fit4[[2]] )
d5%>%
  dplyr::select(Class) %>%
  dplyr::mutate(class = as.factor(Class))%>%
  dplyr::group_by(class)%>%
   dplyr::summarise(n = n()) %>%
   dplyr::mutate(freq = n / sum(n))
```

About 6-7 % are in the low wpPCA -> high sattPCA class. This is the same proportion as in the 3 class solution  


However, I'm not sure I would trust this approach to modelling an LPA.  This is because it relies on a latent profile of a latent variable. I don't have enough background in LPA to assess whether that isn't bat-shit crazy. 



## Just the code


```{r code_folding = FALSE, eval=FALSE}

# libraries
library("tidyLPA") # latent profile analysis
library("here") # file organisation
library("tidyverse") # data wrangling

# import data: note -- you'd need to designate the paths by folders in your library
# my path is : /Users/jbul176/GIT/notes/colleagues/marc/"KDat2403.csv
# here::here() will take you to the root directory and then you specify a location from there

dat<-as_tibble(readr::read_csv(here::here("colleagues", "marc", "KDat2403.csv"), col_names = FALSE))


# Rename colums  
df <- dat %>%
dplyr::rename( "id" = "X1",
       "anx" = "X2",
        "cdct" = "X3", 
       "sh"= "X4",
       "stht" = "X5",
       "satt" = "X6",
       "wb"= "X7",
       "rads"= "X8" )


# LPA without the binary variables


p <- 7 # max number of classes
fit <-  df2 %>%
  dplyr::select(anx_s,
                cdct_s,
                wb_s,
                rads_s) %>%
  tidyLPA::single_imputation() %>%
  tidyLPA::estimate_profiles(3:p)


# Get fit statistics

tidyLPA::get_fit( fit )


# Graph all results and create labels for the graph 

tidyLPA::plot_profiles( fit4, add_line = TRUE ) + 
  ylab("PCA units") + labs(title = "LPA of PCA: no warrenties")


# Plot only the 4-class solution
tidyLPA::plot_profiles( fit [[2]] , add_line = TRUE) ## note we are counting up from three, hence `[[2]]`

# Create table for proportion of sample in classes
df_2 <- tidyLPA::get_data(fit[[2]]) ## note we are counting up from three, hence `[[2]]`
df_2 %>%
  dplyr::select(Class) %>%
  dplyr::mutate(class = as.factor(Class)) %>%
  dplyr::group_by(class) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(freq = n / sum(n))


# notes: the PCA approach above is "experimental" and possibly flawed. However, it suggest an alternative approach to the problem, involving data-dimension reduction. 

# To do: create better tables 

```



## What predicts suicide attempt?


#### Paradoxically, 1  x SD increase in well being predicts suicide .41 greater odds of suicidal attempt.

```{r}
sjPlot::tab_model(
  glm(as.integer(satt) ~ wb_s, data = df2, family = "binomial")
  )
```


####  Suicidal thought predicts 52.17 x greater odds for suicide attempt

```{r}
sjPlot::tab_model(
  glm(as.integer(satt) ~ stht, data = df2, family = "binomial")
  )
```

####  1 x SD of anxiety predicts 2 x greater odds of suicide attempt


```{r}
sjPlot::tab_model(
  glm(as.integer(satt) ~ anx_s, data = df2, family = "binomial")
  )
```


#### Self harm predicts 28.11 times higher chance of suicide attempt!

```{r}
sjPlot::tab_model(
  glm(as.integer(satt) ~ sh, data = df2, family = "binomial")
  )
```


#### 1 sd increase in conduct problem predicts 2 x greater odds of suicide attempt

```{r}
sjPlot::tab_model(
  glm(as.integer(satt) ~ cdct_s, data = df2, family = "binomial")
  )
```


### Depression  4.78 x greater odds of suicide attempt

```{r}
sjPlot::tab_model(
  glm(as.integer(satt) ~ rads_s, data = df2, family = "binomial")
  )
```


#### To do: need a DAG and causal analysis




